---
title: "Software overview"
editor: visual
code-copy: true
---

Read this and complete all instructions in the 'action' boxes during your lab section. Your TA will walk you through the activity and help to troubleshoot issues and answer any questions along the way.

## Prerequisites

<!--# TA ACTION: ensure everyone has working installs, a github account, and access to the org before going ahead. You may need to send github org invites to students; prepare a list of email addresses for this purpose in advance, just in case. There may be time to allow a few minutes for anyone who needs the time to complete installs, but if need be you can pair students. -->

To complete the activity you'll need to:

-   have all of the software listed on the course technology page installed;

-   find (or create) your [GitHub](https://github.com/) account credentials (if you are creating an account for the first time, see [advice on choosing a username](https://happygitwithr.com/github-acct.html#username-advice)).

::: callout-important
## Action

**Preparations:**

1.  Log in to your GitHub account.
2.  Accept the invitation to join the PSTAT197 GitHub organization. Check your UCSB email for the invitation. If you have not received one, request it from a TA or ULA.
3.  Open your GitHub client.
4.  Open a new session in RStudio.
5.  Create a class folder for PSTAT197 somewhere on your machine, *e.g.,* `~/documents/pstat197`.
:::

## RStudio projects

First we'll get acquainted with the basic functionality of the RStudio IDE and the use of projects as a means of organizing files. If you've already used RStudio, great -- this will still serve to introduce you to how we'll use RStudio projects in this class.

### RStudio Setup

<!--# TA ACTION: take a minute or two to review the IDE layout in RStudio. point out the bulletted items and say a word or two about each one. -->

Your TA will briefly review the (default) layout of the RStudio IDE. You should be able to identify/find the following:

-   console

-   terminal

-   file navigator

-   environment

-   history

We'll use several R packages throughout the quarter. Some of these we will install on the go, but we can install several that we'll rely on now.

::: callout-important
## Action

**Install packages**

Navigate to the console and copy-paste the following commands. You only need to do this once. This will take a minute or two to complete.
:::

```{r, eval=F}
# package install list 
url <- 'https://raw.githubusercontent.com/pstat197/pstat197a/main/materials/scripts/package-installs.R'
source(url)

# clear environment
rm(list = ls())
```

### Create a local project

*Projects* are a means of keeping your work organized. When you create a project in a directory on your local machine, RStudio keeps track of project metadata, history, and the working environment so that every time you open the project you see whatever you had open when you last closed it.

::: callout-important
## Action

**Create a new project**:

-   Select *File \> New project*

-   Create the project in a new directory as a subdirectory of your class folder

-   Name it something like `example-project`

*Comment:* when naming files it's good practice to avoid spaces, special characters, and the like. A naming convention we try to follow: choose a descriptive name comprising 1-3 words or common abbreviations separated by hyphens.
:::

Take a moment to observe the file navigator. It should consist of a single `example-project.Rproj` file.

### Add content

We may as well populate the project with a few files -- so let's add a dataset and write a short script, as if we're just starting a data analysis.

::: callout-important
## Action

**Retrieve data and store a local copy**

-   Open a new script: *File \> New File \> R Script*

-   In the navigator, create a folder called `data` and a folder called `scripts`

-   Copy and paste the code chunk below into your script.

-   Execute once, then save in the `scripts` folder as `data-retrieval.R` and close
:::

```{r, eval = F}
library(tidyverse)

# retrieve pollution data
url <- 'https://raw.githubusercontent.com/pstat197/pstat197a/main/materials/labs/lab1-setup/data/pollution.csv'
pollution <- read_csv(url)

# write as csv to file
write_csv(pollution, file = 'data/pollution.csv')

# clear environment
rm(list = ls())
```

<!--# TA COMMENT: you might pause here and mention a few things. first, we've used a script (not an rmarkdown file) because we only have R codes and no text content (many students acquire the bad habit of working in rmarkdown when writing codes). second, storing a script like this might seem silly but it does establish a record of the data source -->

Next, we'll do a simple regression analysis.

::: callout-important
## Action

**Create a script**

-   Create a new script as before

-   Copy-paste the code chunk below into your script

-   Execute once and examine the results

-   Save in the `scripts` folder as `slr-analysis.R`
:::

```{r, message = F, warning = F}
library(tidyverse)

# load data
pollution <- read_csv('data/pollution.csv')

# examine scatterplot with SLR fit
ggplot(pollution,
       aes(x = log(SO2), y = Mort)) +
  geom_point() +
  geom_smooth(method = 'lm')

# compute SLR fit
fit <- lm(Mort ~ log(SO2), data = pollution)
broom::tidy(fit)

# interpret
fit_ci <- confint(fit, parm = 'log(SO2)')*log(1.2)

paste('With 95% confidence, every 20% increase in sulfur dioxide pollution is associated with an increase in two-year mortality rate between', 
      round(fit_ci[1], 2), 
      'and', 
      round(fit_ci[2], 2), 
      'per 100k', sep = ' ') %>% 
  print()
```

<!--# TA COMMENT: show everyone what their directories should look like at completion. mention that the project treats the project directory as the working directory, so scripts stored anywhere in the project still parse directory paths relative to the project directory, NOT the file location. -->

Congrats on your first project! You can close the RStudio session now.

We'll be using projects structured much like what you just set up, but with one catch: we'll link up our RStudio projects with shared *repositories* so that we can all collaborate on the same set of project files.

## GitHub repositories

<!--# TA COMMENT: you could either talk through this section or have students read it and clone the repo themselves. either way, show the use of the client, the local copy, and the remote location (on github.com). the goal here is for students to understand the distinction between local and remote and that the client provides a means of communication -->

### Git and GitHub

At some point in time -- possibly quite recently -- you had to install Git on your local machine, as well as create a GitHub account. So, Git and GitHub are two different things.

**Git is version control software** that enables you to systematically track and control file changes within a *repository* -- a collection of files possibly with some directory structure. (The definition of 'repository' is simply 'storage place'.)

**GitHub is an online platform** for hosting repositories remotely. Anyone with access to a repository can make changes to files in the repository, and this enables multiple people to collaborate on code.

### local \<\> remote

Usually remote repositories are not updated directly because contributors need to execute codes to test their changes and the remote server that hosts the repository is not equipped to do this.

Instead, contributors will prepare changes on their own machine where they can test them, and then update the remote repository once their changes are complete.

This process of implementing file changes in a repository involves communicating information between *local* and *remote* locations. For this purpose a local copy of the remote repository is needed.

### Cloning a repository

In Git lingo, a *clone* is a local copy of a remote repository. Creating a clone copies files and establishes the link between local and remote repositories so that changes can be sent to and received from the remote repository. ***You only need to create a clone once.***

To clone a repository, all one needs is:

-   the remote location URL;

-   the local destination where the clone will be created;

-   permission from the repository owner, if private.

I've created a 'sandbox' repository (for playing in) at <https://github.com/pstat197/sandbox>. It's a public repository, so no special permissions are needed. Follow the link to see the repository contents in your browser.

::: callout-important
## Action

**Clone the sandbox repository:**

1.  Open your GitHub client (GitKracken or GitHub Desktop or similar) and ensure you are logged in to your GitHub account.
2.  Look for a 'Clone Repo' menu item or similar and simply input the URL and the place you'd like to clone it; proceed through any prompts.
3.  Check your file navigator to confirm that the repository files were copied.
:::

<!--# TA COMMENT: remind students we'll use this repository next class meeting, so DON'T delete the clone -->

An alternative possibility is to create the clone using a terminal command. In the terminal, navigate to the desired destination, and input:

``` bash
git clone https://github.com/pstat197/sandbox
```

::: callout-note
## Remarks

**On terminal commands:**

-   It's recommended to manage Git actions through a visual client, as it's much easier to see and understand what's happening.

-   However, if you know exactly what you're doing, executing simple actions via Git bash in the terminal can be more efficient at times.

-   For example, you can keep a terminal open in RStudio and manage your repository workflow from there, without having to toggle between environments.

-   Try experimenting with terminal commands from RStudio after you have a little experience with basic Git actions.
:::

## Checklist

Have you completed all of the activity action items?

1.  Install software: R, RStudio, Git, and a GitHub client
2.  Create a GitHub account and join the [course GitHub org](https://github.com/pstat197/)
3.  Install R packages that will be used frequently
4.  Create a local project in RStudio
5.  Clone the sandbox repo <https://github.com/pstat197/sandbox>

# 
