---
title: "Training neural networks"
editor: visual
code-copy: true
execute:
  message: false
  warning: false
  echo: true
  cache: true
---

## Pre-lab activity

Please complete this activity before arriving at your section meeting. It should only take you about 15 minutes.

### Python installation

We will use packages that interface to python from R to train neural networks in this lab. For this a current python installation is needed.

Select the download appropriate for your operating system [here](https://www.python.org/downloads/) and follow installation instructions.

### Keras and Tensorflow

Open RStudio and execute the following commands in the console. You should have already installed `keras` with other packages in the first lab; however, if `library(keras)` returns an error execute `install.packages('keras')` and then try again. This will install Tensorflow (for python) in a manner suitable for use in R, along with a few other packages.

```{r}
#| eval: false
library(keras)
install_keras()
```

To confirm the installation worked, try:

```{r}
library(tensorflow)
tf$constant('Hello world')
```

You may see a long message related to CUDA libraries in addition to the output shown above, but if you see this output at the end, your installation was successful.

If you did not see the expected output, try configuring a virtual environment for the installation explicitly as shown [here](https://tensorflow.rstudio.com/install/#installation).

If you are unable to troubleshoot after a short period of time, partner with a classmate for the lab activity and then ask for help from course staff.

## Lab activity
