---
title: "Results from claims assignment"
subtitle: "PSTAT197A/CMPSC190DD Fall 2022"
author: "Trevor Ruiz"
institute: 'UCSB'
bibliography: refs.bib
format: 
  revealjs:
    incremental: true
    # footer: 'PSTAT197A/CMPSC190DD Fall 2022'
    # logo: 'img/ucsbds_hex.png'
    fig-width: 6
    fig-height: 4
    fig-align: 'left'
    slide-number: true
code-copy: true
execute:
  message: false
  warning: false
  echo: true
  cache: true
---

## Today's agenda

1.  Review results of claims assignment
2.  Discuss vignette guidelines
3.  Remaining time to organize in groups

## Preface

My goal here is to use the results of your work on the last assignment to learn as a group what worked well and what didn't.

. . .

As a disclaimer, many groups did not successfully generate predictions according to instructions. This is okay.

. . .

My goal is ***not*** to judge anyone based on results or having completed the assignment.

## Accessing claims predictions

Results from the last assignment are now available online.

```{r}
library(tidyverse)

github_url <- "https://github.com/pstat197/pstat197a/raw/main/materials/slides/data/f22-claims-evals.RData"

load(url(github_url))

evals %>% head()
```

## Checking errors: example

```{r}
evals %>%
  filter(group == 4) %>%
  pull(eval)
```

## Accuracies {.scrollable}

```{r}
accuracies <- evals %>%
  rowwise() %>%
  filter(is_tibble(eval)) %>%
  ungroup() %>%
  unnest(eval)

accuracies %>% head(8) %>% knitr::kable()
```

## Comparisons

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

theme_set(theme(text = element_text(size = 20)))
accuracies %>%
  ggplot(aes(x = factor(group), y = .estimate)) +
  geom_point(aes(alpha = n), size = 4) +
  facet_wrap(~class*.metric, nrow = 2) +
  labs(x = 'group', y = '')
```

## Vignette guidelines

Let's review as a group.
